\documentclass{beamer}
\usepackage{amsthm}
\usepackage[utf8]{inputenc}
\usetheme{madrid}
\usepackage{../latex_style/packages}
\usepackage{../latex_style/notations}
\usepackage{outlines}

\def\labelenumi{\theenumi}

\setbeamertemplate{itemize items}[default]
\setbeamertemplate{enumerate items}[default]

\newcommand\Fonteight{\fontsize{8}{9.6}\selectfont}

%Information to be included in the title page:
\title{Recitation 1 }
\author{Alex Dong}
\institute{CDS, NYU}
\date{Fall 2020}


\makeatletter
\setbeamertemplate{footline}{
  \leavevmode%
  \hbox{%
  \begin{beamercolorbox}[wd=.4\paperwidth,ht=2.25ex,dp=1ex,center]{author in head/foot}%
    \usebeamerfont{author in head/foot}\insertshortauthor\expandafter\ifblank\expandafter{\beamer@shortinstitute}{}{~~(\insertshortinstitute)}
  \end{beamercolorbox}%
  \begin{beamercolorbox}[wd=.3\paperwidth,ht=2.25ex,dp=1ex,center]{title in head/foot}%
    \usebeamerfont{title in head/foot}\insertshorttitle
  \end{beamercolorbox}%
  \begin{beamercolorbox}[wd=.3\paperwidth,ht=2.25ex,dp=1ex,right]{date in head/foot}%
    \usebeamerfont{date in head/foot}\insertshortdate{}\hspace*{2em}
    \insertframenumber{} / \inserttotalframenumber\hspace*{2ex} 
  \end{beamercolorbox}}%
  \vskip0pt%
}
\makeatother

\begin{document}
%1
\frame{\titlepage} 
%2
\begin{frame}
\frametitle{Announcements}
\end{frame}

%2
\begin{frame}
\frametitle{Why is Linear Algebra important?}
Linear Algebra ...
\begin{itemize}
\item Appears in ALL applied math, including \textit{data science}
\item Is solvable. If you can write it down, you can solve it! \\
    \qquad (not true for other math, e.g differential equations, integration)
\item Is \textit{fundamental} to understanding tools in machine learning

\end{itemize}

Relevant applications we will cover in the class:
\begin{itemize}
\item Linear Regression
\item Principal Component Analysis
\item Gradient Descent
\end{itemize}

\end{frame}
%3
\begin{frame}
\frametitle{Concept Review: Vector Spaces}
\begin{definition}[Vector space]
	    
	    $V$ is a vector space over field $F$ when
	    \Fonteight
		\begin{enumerate}
		\item[1.] Closure under Addition: $x+y \in V$
		\item[2.] Sum is commutative ($x+y = y+x$) and associative $x+(y+z)= (x+y)+z$
		\item[3.] Additive Identity (in $V$): $0\in V$ ($x+0 =x$)
		\item[4.] Additive Inverse: ($\forall x, \exists -x $ s.t $ x+(-x)=0$
		\item[5.] Closure under Scalar Multiplication: $\alpha x \in V$
		\item[6.] Multiplicative Identity (in $F$): $\alpha x \in V$
		\item[7.] Compatibility in Multiplication: $\alpha (\beta x) = (\alpha \beta) x $
		\item[8.] Distributivity: $(\alpha + \beta) x = \alpha x + \beta \cdot \vec{y}
			 \text{ and }
			\alpha  (x + y) = \alpha  x + \alpha \cdot y.$
		\end{enumerate}
\end{definition}
Notice that "vectors" are not explicitly defined. \\

Optional: prove (after class) that $\R^3$ with standard definitions for addition and scalar multiplication is vector space over field $\R$ .

\end{frame}
%4
\begin{frame}
\frametitle{Concept Review: Vector Spaces}
In this class,
\begin{itemize}
\item Our field is always $\R$. ($\C$ is also a field)
\item Standard definitions for vector addition, and scalar multiplication
\item But, V is (usually) $\R^n$, or (sometimes) $\R^{n\times n}$
\end{itemize}


Everything in Linear Algebra is in a vector space.
\end{frame}
\begin{frame}
\frametitle{Concept Review: Subspaces}
\begin{definition}[Subspace]
	A subset $S$ of a vector space $V$ is a \emph{subspace} if it is closed under addition and scalar multiplication.
	\begin{enumerate}
		\item[1.] Closure under Addition: $x+y \in V$
		\item[2.] Closure under Scalar Multiplication: $\alpha x \in V$
	\end{enumerate}
\end{definition}
A subspace is also a vector space!\\
Everything in linear algebra is in a vector space. \\ 
Anything \textit{interesting} in linear algebra is in a subspace.
Subspaces are a recurrent concept throughout this entire course.
\end{frame}
\begin{frame}
\frametitle{Questions 1: Subspaces, Span}

\end{frame}
\begin{frame}



\frametitle{Solutions 1, Linear Independence, Bases, }

\end{frame}



\end{document}