\section{Recitation 1}
\begin{enumerate}
\item Recall that $\RR^2=\{(x,y) : x,y\in\RR\}$ can be thought
  of as the $xy$-plane.  Consider the two vectors $v=(9,7)$ and
  $w=(-8,12)$.  Describe the following sets geometrically.  Which are
  subspaces of $\RR^2$?
  \begin{enumerate}
  \item $\Span(v)$
  \item $\Span(v,w)$
  \item $\Span(v)\cup\Span(w)$, that is, the vectors in $\Span(v)$ or
    $\Span(w)$
  \item $\Span(v)\cap\Span(w)$, that is, the vectors in both $\Span(v)$ and
    $\Span(w)$
  \item $\{(1-t)v + tw : t\in[0,1]\}$
  \item $\{(1-t)v+tw : t\in\RR\}$
  \item $\{\alpha v + \beta w : \alpha,\beta\geq 0\}$
  \item $\Span(v,w,x)$ where $x=(0,5)$.
  \item $\{(a,b)\in\RR^2 : a^2+b^2\leq 25\}$
  \item $\{(a,a)\in\RR^2 : a\in\RR\}$
  \item $\{(a,a^2)\in\RR^2 : a\in\RR\}$
  \item $\{(a,1)\in\RR^2 : a\in\RR\}$
  \end{enumerate}
\begin{solution}
\item[]\Sol Below recall that $W$ is a subspace of $\RR^n$ if it is
  nonempty and satisfies
  \begin{itemize}
  \item $ax\in W$ for all $a\in\RR$ and $x\in W$
  \item $x+y\in W$ for all $x,y\in W$
  \end{itemize}
  \begin{enumerate}
  \item Line through $v$.  It is a subspace.
  \item All of $\RR^2$. It is a subspace.
  \item 2 lines through $v$ and $w$.  Not a subspace (doesn't contain
    $v+w$).
  \item Only the $0$ vector. It is a subspace.
  \item Line segment between $v$ and $w$.  Not a subspace (doesn't
    contain $0$).
  \item Line through $v$ and $w$.  Not a subspace (doesn't contain
    $0$).
  \item Infinite wedge between $v$ and $w$.  Not a subspace (doesn't contain
    $-v$).
  \item All of $\RR^2$.  It is a subspace.
  \item All points within a circle of radius 5 centered at the
    origin.  Not a subspace (doesn't contain $2\cdot(5,0)$).
  \item Diagonal line through the origin.  It is a subspace.
  \item Parabola.  Not a subspace (doesn't contain $2\cdot(1,1)$).
  \item Horizontal line with $y$-coordinate equal to $1$.  Not a
    subspace (doesn't contain $0$).
  \end{enumerate}
\end{solution}
\item Let $v_1,v_2,v_3,v_4\in\RR^3$ be distinct and define $C_1=\{v_1,v_2\}$ 
  and $C_2=\{v_3,v_4\}$.  If $C_1$ and $C_2$ are both linearly
  independent, what are the possible values for
  $\dim(\Span(v_1,v_2,v_3,v_4))$? No proof necessary.
\begin{solution}
\item[]\Sol Either $C_1\subset\Span(C_2)$ and the dimension is $2$, or
  the dimension is $3$.  We will discuss dimension and justify this later in the class.
\end{solution}
\item Let $B=v_1,\ldots,v_n$ be a basis for $\RR^n$.  Show that
  for any $x\in\RR^n$ there exists a unique $\alpha\in\RR^n$ such that
  $$x = \alpha_1v_1 + \cdots + \alpha_nv_n.$$
\begin{solution}
\item[]\Sol Existence follows immediately from the definition of
  basis, since it implies that $\Span(v_1,\ldots,v_n)=\RR^n$.  For uniqueness,
  suppose
  $$x = \alpha_1v_1+\cdots+\alpha_nv_n =
  \beta_1v_1+\cdots+\beta_nv_n$$
  for $\alpha,\beta\in\RR^n$.  Then we have
  $$0 = (\alpha_1-\beta_1)v_1+\cdots+(\alpha_n-\beta_n)v_n.$$
  By linear independence, we must have $\alpha_i=\beta_i$ for all $i$.
\end{solution}
\item True or False: If $B=\{v_1,\ldots,v_n\}$ is a basis for $\RR^n$,
  and $W$ is a subspace of $\RR^n$, then some subset of $B$ is a basis
  for $W$.
\begin{solution}
\item[]\Sol False. Consider $B=\{(1,0),(0,1)\}$ and $W=\Span((1,1))$.
\end{solution}
\item Suppose $v_1,\ldots,v_m\in\RR^n$ are linearly dependent. Prove
  that if
  $x\in\Span(v_1,v_2,\ldots,v_m)$ then there are infinitely many
  $\alpha\in\RR^m$ with
  $$x = \alpha_1v_1 + \cdots + \alpha_m v_m.$$
\begin{solution}
\item[]\Sol
  \begin{proof}
    By assumption $x\in\Span(v_1,\ldots,v_m)$ so
    $$x=a_1v_1+\cdots+a_mv_m$$
    for some $a_i\in\RR$.  Since $v_1,\ldots,v_m$ are linearly
    dependent, there are $c_1,\ldots,c_m\in\RR$ such that
    $$c_1v_1+\cdots+c_mv_m = 0$$
    where not all $c_i=0$.  Then we have
    $$x = a_1v_1+\cdots+a_mv_m + r(c_1v_1+\cdots+c_mv_m)
    = (a_1+rc_1)v_1+\cdots + (a_m+rc_m)v_m$$
    for all $r\in\RR$.  This gives infinitely many distinct $\alpha$
    where $\alpha_i=a_i+rc_i$ for $r\in\RR$.
  \end{proof}
\end{solution}
\item Find (if they exist) the (global) maximizers and minimizers of the
  following function $f:\RR\to\RR$ defined by
  $$f(x) = |x-1|+|x-3|+|x-10|.$$
\begin{solution}
\item[]\Sol There is no maximizer since $f(x)\to\infty$ as
  $x\to\infty$.  Where $f'$ exists, it is negative when $x<3$ and
  positive when $x>3$ so $3$ is the minimizer. [Technically, we are
    applying the mean value theorem to show $f$ is descending on
    $(-\infty,1)$ and $(1,3)$ and ascending on $(3,10)$ and $(10,\infty)$.]
\end{solution}
\end{enumerate}
